Unshuffling Permutations

### 2015-07-02 article.tex
Partie algebre (modifications)

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
Let use denote by $\QQ[S]$ the linear span of all permutations. We shall
define on this space a kind of shuffle product. For this purpose, we
first define a linear coproduct $\Delta$ on $\QQ[S]$ satisfying for
any permutation $\pi$,
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \uplus P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
\medskip

@@

>>>>

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
Let use denote by $\QQ[S]$ the linear span of all permutations. We shall
define on this space a kind of shuffle product. For this purpose, we
first define a linear coproduct $\Delta$ on $\QQ[S]$ satisfying for
any permutation $\pi$,
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \uplus P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract two disjoint subwords of the permutation $1432$
so that the letters of these two subwords have the same relative order
as the letters of the permutations $1$ and $132$.
\medskip

@@

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. There are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the {\em shifted shuffle} and the
{\em convolution product}. Our shuffle product is different from these
last two.
\medskip
